---
title: "Mold project Census data cleaning"
author: "Jenna Tipaldo"
date: "2025-07-30"
output: html_document
---

# Load Packages

```{r load}
library(data.table)
library(tidyverse)
library(lubridate) #for dates
library(zoo) # for yearmon()
library(stats) #for ts()
library(TSstudio)
library(haven)
library(stargazer)
library(sjPlot) #model tables
library(tidycensus)
census_api_key("") #REPLACE WITH YOUR OWN KEY
```

#Load tract file unclipped (with water) from NYC DCP & clean
2020 tracts Latest Release: 25C, Date of Data: August 2025
https://www.nyc.gov/content/planning/pages/resources/datasets/census-tracts 
```{r}
tracts20<-st_read("nyctracts2020water/nyct2020wi.shp")

tracts20<- tracts20%>%
  select(GEOID, BoroName, NTA2020, Shape_Area)
```


# Census: total pop from tidycensus & join to shapefile
```{r}
totpop20<-get_decennial(
  geography = "tract",
  state = "NY",
  variables = "P1_001N",
  year = 2020,
  #sumfile = "dhc",
  geometry = F
  )

summary(totpop20$value)

totpop20<- totpop20%>%
  rename(totpop=value)%>%
  select(-variable, -NAME)


tracts20 <- left_join(tracts20, totpop20, by="GEOID")
```

Census API tips: use NHGIS to search and find source tables, then use the NHGIS codebook or this site to determine what the variables represent when you do the API pull: https://api.census.gov/data.html

# Read in census data and recode to BCT (boro census tract for merge)
Value | Borough name | FIPS State County code
1 Manhattan 36061
2 Bronx 36005
3 Brooklyn 36047
4 Queens 36081
5 Staten Island 36085
```{r}
# ALTERNATIVE: NHGIS data
#nhgis_blocks<-read.csv("nhgis0023_ds258_2020_block.csv")

#List tables of interest (see above, can use NHGIS to help find tables)
tables <- c("H4") #, "P15", "H10"

# Loop over each table and bind the results into one data frame
tracts20_variables <- map_dfr(tables, ~get_decennial(
  geography = "tract",
  state = "NY",
  table = .x,
  year = 2020,
  sumfile = "dhc",
  geometry = F
  ))

head(tracts20_variables)

tracts20_variables_wide <- tracts20_variables %>%
  filter(substr(GEOID, 1,5)%in%c("36061","36005","36047","36081","36085"))%>%
  dplyr::select(GEOID, NAME, variable, value) %>%
  pivot_wider(names_from = variable, values_from = value)%>%
  mutate(BCT2020 = case_when(
    substr(GEOID, 1,5) == "36061" ~ paste0("1",substr(GEOID, 6,11)),
    substr(GEOID, 1,5) == "36005" ~ paste0("2",substr(GEOID, 6,11)),
    substr(GEOID, 1,5) == "36047" ~ paste0("3",substr(GEOID, 6,11)),
    substr(GEOID, 1,5) == "36081" ~ paste0("4",substr(GEOID, 6,11)),
    substr(GEOID, 1,5) == "36085" ~ paste0("5",substr(GEOID, 6,11))
  ),
  BCT2020 = as.integer(BCT2020))

tracts20_variables_wide <- tracts20_variables_wide %>%
  mutate(check = if_else(H4_001N == H4_002N+H4_003N+H4_004N, 1, 0))%>%
  rename(
    housingunits_total=H4_001N,
    housingunits_ownedmortgage=H4_002N,
    housingunits_ownedclear=H4_003N,
    housingunits_renters=H4_004N,
  )

table(tracts20_variables_wide$check, useNA = "always")

tracts20_variables_wide<- tracts20_variables_wide%>%
  select(-NAME, -check)

tracts20 <- left_join(tracts20, tracts20_variables_wide, by="GEOID")
```


#additional census info (still needs cleaning based on geography changes)
```{r}
#List tables of interest (see above, can use NHGIS to help find tables)
tables <- c("H10","H12") #, "P15", "H10"

# Loop over each table and bind the results into one data frame
tracts20_additional <- map_dfr(tables, ~get_decennial(
  geography = "tract",
  state = "NY",
  table = .x,
  year = 2020,
  sumfile = "dhc",
  geometry = F
  ))

codebook <- load_variables(2020, "acs5") #add /subject
view(codebook)

head(tracts20_additional)

tracts_ACS_wide <- tracts20_additional %>%
  filter(substr(GEOID, 1,5)%in%c("36061","36005","36047","36081","36085"))%>%
  dplyr::select(GEOID, NAME, variable,source,estimate, moe) %>%
  pivot_wider(id_cols = c(GEOID), names_from = c(variable, source), values_from = c(estimate, moe))%>%
  mutate(BCT2020 = case_when(
    substr(GEOID, 1,5) == "36061" ~ paste0("1",substr(GEOID, 6,11)),
    substr(GEOID, 1,5) == "36005" ~ paste0("2",substr(GEOID, 6,11)),
    substr(GEOID, 1,5) == "36047" ~ paste0("3",substr(GEOID, 6,11)),
    substr(GEOID, 1,5) == "36081" ~ paste0("4",substr(GEOID, 6,11)),
    substr(GEOID, 1,5) == "36085" ~ paste0("5",substr(GEOID, 6,11))
  ),
  BCT2020 = as.integer(BCT2020))

# years <- c(2013, 2018, 2023) #if can figure out continuity
years <- c(2020, 2021)
tracts20only_additional <- map_dfr(years, ~get_acs(
  geography = "tract",
  state = "NY",
  survey = "acs5",
  #table = .x,
  variables = c(medianincome = "B19013_001", englishlow_foreignborn="B16005I_011", englishlow_native="B16005I_006",english_total="B16005I_001",
                race_NHwhite="B03002_003",race_NHblack="B03002_004",race_hispanic="B03002_012", race_total="B03002_001"),
  year = .x,
  geometry = F
  )%>% 
    mutate(source=.x) #adds column for the ID
  )


tracts20only_ACS_wide <- tracts20only_additional %>%
  filter(substr(GEOID, 1,5)%in%c("36061","36005","36047","36081","36085"))%>%
  dplyr::select(GEOID, NAME, variable,source,estimate, moe) %>%
  pivot_wider(id_cols = c(GEOID), names_from = c(variable, source), values_from = c(estimate, moe))

summary(tracts20only_ACS_wide$estimate_medianincome_2020)

tracts20only_ACS_wide_recode <- tracts20only_ACS_wide %>%
  mutate(lowenglish20 = (estimate_englishlow_foreignborn_2020+estimate_englishlow_native_2020)/estimate_english_total_2020 *100,
         NHblack20 = estimate_race_NHblack_2020 / estimate_race_total_2020 *100,
         medinc20 = estimate_medianincome_2020)%>%
  select(GEOID, medinc20, NHblack20, lowenglish20)

tracts20<- left_join(tracts20,tracts20only_ACS_wide_recode, by="GEOID")
```



# save final shapefile with Census info
```{r}

write.csv(tracts20, "nyc_tracts_2020_info.csv")

tracts20_namescleaned<- tracts20%>%
  select(-housingunits_ownedmortgage, -housingunits_ownedclear,
         -Shape_Area)%>%
  rename(
    HU_tot = housingunits_total,
    HU_rent= housingunits_renters
  )

st_write(tracts20_namescleaned, "nyc_tracts_2020_info.shp")

```


# read in and clean HUD NYCHA data
```{r}
publichousing<-read.csv("publichousing_2020tracts_in2022.csv")

publichousing_recode<- publichousing%>%
  rename(housingunits_publichousing = "Subsidized.units.available",
         occupied_pct = "X..Occupied",
         GEOID = "Code")%>%
  mutate(GEOID = as.character(GEOID),
         #for missing values (-4 is suppressed, -5 is lack of sufficient reporting, impute mean prop occupied)
         occupied_prop = if_else(occupied_pct<0 , mean(occupied_pct)/100, occupied_pct/100),
         housingunits_pubhousingoccupied = round(housingunits_publichousing * occupied_prop )
         )#%>%
  select(GEOID, housingunits_pubhousingoccupied)

sum(publichousing_recode$housingunits_pubhousingoccupied)

```

# join HUD data to census tract data and calculate units 
```{r}
tracts_housingunits <- left_join(tracts20, publichousing_recode, by="GEOID")

summary(tracts_housingunits$housingunits_publichousing)

tracts_housingunits <- tracts_housingunits %>%
  #zero out NA tracts
  mutate(
    #flag for any PH in tract
    anypublichousing_tract = if_else(housingunits_pubhousingoccupied>0,1,0),
    #replace NAs with 0s
    housingunits_pubhousingoccupied = if_else(is.na(housingunits_pubhousingoccupied), 0, housingunits_pubhousingoccupied),
    #replaces 5 tracts where PH units from HUD for 2022 were higher than 2020 total renters
         housingunits_pubhousingoccupied = if_else(housingunits_pubhousingoccupied>housingunits_renters, housingunits_renters, housingunits_pubhousingoccupied), 
    rentalunits_minusPH = housingunits_renters - housingunits_pubhousingoccupied
         )
```

# save final shapefile & csv with public housing units

```{r}
write.csv(tracts_housingunits, "nyc_tracts_2020_infoHU.csv")

tracts20_namescleaned_HU<- tracts_housingunits%>%
  select(-housingunits_ownedmortgage, -housingunits_ownedclear, 
         -Shape_Area, -housingunits_pubhousingoccupied, -anypublichousing_tract)%>%
  rename(
    HU_tot = housingunits_total,
    HU_rent= housingunits_renters,
    HUrentPH = rentalunits_minusPH
  )

st_write(tracts20_namescleaned_HU, "nyc_tracts_2020_infoHU.shp")


```

#shapefile with population
```{r}

summary(tracts_housingunits$totpop)

quantile(tracts_housingunits$totpop, probs = .05)

quantile(tracts_housingunits$housingunits_total, probs = .05)

withpopulation<-tracts_housingunits%>%
  filter(totpop>=500 | housingunits_total>=100)%>%
  select(-housingunits_ownedmortgage, -housingunits_ownedclear, 
         -Shape_Area, -housingunits_pubhousingoccupied, -anypublichousing_tract)%>%
  rename(
    HU_tot = housingunits_total,
    HU_rent= housingunits_renters,
    HUrentPH = rentalunits_minusPH
  )

excluded<-tracts_housingunits%>%
  filter(totpop<500 | housingunits_total<100)


st_write(withpopulation, "nyc_tracts_2020_infoHU_withpop.shp")
```


