---
title: "Mold_Figures"
author: "Jenna Tipaldo"
date: "2026-7-24"
output: html_document
---

# Load Packages

```{r load}
library(data.table)
library(tidyverse)
library(lubridate) #for dates
library(zoo) # for yearmon()
library(stats) #for ts()
library(TSstudio)
library(haven)
library(lme4) #for multilevel modeling
library(stargazer)
library(sjPlot) #model tables
library(sjlabelled) #for labels in plots
library(sf) #for spatial shapefiles
library(beepr) #to play a sound when it's done running
```

#Load data
```{r}
#file of overall 311 calls for any issue and for mold
dailycomparison<- read.csv("mold_dailycomparison.csv")
#mold complaints
complaintsrecoded <- read_csv("complaintsrecoded.csv")
```

#Process data weekly
```{r}
#weekly
date1 <- as.Date("2010-01-01")

weeklycomparison <- dailycomparison%>%
  mutate(weeknum = as.numeric(floor(difftime(day, date1, units = "weeks"))+1)
         # Week_chr = str_pad(Week, width = 2, side = "left", pad = "0"),
         # year_isoweek = paste0(Year,"-",Week_chr)
         )%>%
  #for each week
  group_by(weeknum)%>%
  summarize(weekstart = min(day),
            week_end = max(day),
            weeklyn_mold = sum(dailyn_mold, na.rm = T), #
            weeklyn_overall = sum(dailyn_overall, na.rm = T), #
            )%>%
  ungroup()

weeklycomparison <-weeklycomparison %>%
  mutate(week_end = as.Date(week_end))

```



#Figure 1
```{r}
multiplier<-100 #to scale y-axis for the mold complaints

figure1<-ggplot(weeklycomparison, aes(x=week_end))+
  geom_line(aes(y=weeklyn_overall), color="black")+
  geom_line(aes(y=weeklyn_mold*multiplier), color="darkgreen", alpha=0.7)+
  scale_y_continuous(
    name = "311 complaints for any issue (black)",
    labels = scales::comma, #add comma
    sec.axis = sec_axis(~ . / multiplier, name = "311 complaints for mold (green)",) # Inverse transformation for labels
    
  )+
  scale_x_date(date_breaks= "1 year", date_labels = "%Y")+
  labs(x="Week since 1/1/10")+
  theme_classic(base_size = 24)+
  geom_vline(aes(xintercept =as.Date("2011-08-21")), color="grey30",linetype=2, alpha=0.5)+
  annotate("text",x=as.Date("2011-08-21"), y=158, color="grey30", label="Irene", size=6)+
  geom_vline(aes(xintercept =as.Date("2012-10-29")), color="red", linetype=2,alpha=0.5)+
  annotate("text",x=as.Date("2012-10-29"), y=158, color="red", label="Sandy",size=6)+
  geom_vline(aes(xintercept =as.Date("2019-01-01")), color="black", linetype=2,alpha=0.5)+
  annotate("text",x=as.Date("2019-01-01"), y=158, color="black", label="Local Law 55",size=6)+
  geom_vline(aes(xintercept =as.Date("2021-09-01")), color="blue", linetype=2, alpha=0.5)+
  annotate("text",x=as.Date("2021-09-01"), y=158, color="blue", label="Ida",size=6)+
  geom_vline(aes(xintercept =as.Date("2023-09-29")), color="darkred", linetype=2, alpha=0.5)+
  annotate("text",x=as.Date("2023-09-29"), y=158, color="darkred", label="Ophelia",size=6)

figure1
ggsave("mold_figure1.png", figure1, width = 16, height = 8, dpi = 600)
```

#Figure 2
##calculate overall and after storms, then join
```{r}
tract_overall<-complaintsrecoded%>%
  filter(!is.na(BCT2020))%>%
  group_by(BCT2020)%>%
  summarize(n_ovrall = n())%>%
  ungroup()

test<-complaintsrecoded%>%
  filter(!is.na(BCT2020))%>%
  mutate(storm_period90 = case_when(
          day > as.Date("2011-08-21") & day <= (as.Date("2011-08-21")+90) ~ "Iren",
          day > as.Date("2012-10-29") & day <= (as.Date("2012-10-29")+90) ~ "Sand",
          day > as.Date("2021-09-01") & day <= (as.Date("2021-09-01")+90) ~ "Ida",
         day > as.Date("2023-09-29") & day <= (as.Date("2023-09-29")+90) ~ "Ophe"))%>%
  group_by(BCT2020, storm_period90)%>%
  summarize(n = n())%>%
  ungroup()

tracts_stormperiods90<-complaintsrecoded%>%
  filter(!is.na(BCT2020))%>%
  mutate(storm_period90 = case_when(
          day > as.Date("2011-08-21") & day <= (as.Date("2011-08-21")+90) ~ "Iren",
          day > as.Date("2012-10-29") & day <= (as.Date("2012-10-29")+90) ~ "Sand",
          day > as.Date("2021-09-01") & day <= (as.Date("2021-09-01")+90) ~ "Ida",
         day > as.Date("2023-09-29") & day <= (as.Date("2023-09-29")+90) ~ "Ophe"))%>%
  group_by(BCT2020, storm_period90)%>%
  summarize(n = n())%>%
  ungroup()%>%
  complete(BCT2020, storm_period90)%>%
  mutate(n= if_else(is.na(n),0,n))%>%
  filter(!is.na(storm_period90))%>%
  pivot_wider(id_cols = BCT2020, 
              names_from = storm_period90, 
              values_from = n,
              names_glue = "{storm_period90}3m" # Adds "_suffix" to each new column name)
              )

tracts_stormperiods180<-complaintsrecoded%>%
  filter(!is.na(BCT2020))%>%
  mutate(storm_period180 = case_when(
          day > as.Date("2011-08-21") & day <= (as.Date("2011-08-21")+180) ~ "Iren",
          day > as.Date("2012-10-29") & day <= (as.Date("2012-10-29")+180) ~ "Sand",
          day > as.Date("2021-09-01") & day <= (as.Date("2021-09-01")+180) ~ "Ida",
         day > as.Date("2023-09-29") & day <= (as.Date("2023-09-29")+180) ~ "Ophe"))%>%
  group_by(BCT2020, storm_period180)%>%
  summarize(n = n())%>%
  ungroup()%>%
  complete(BCT2020, storm_period180)%>%
  mutate(n= if_else(is.na(n),0,n))%>%
  filter(!is.na(storm_period180))%>%
  pivot_wider(id_cols = BCT2020, 
              names_from = storm_period180, 
              values_from = n,
              names_glue = "{storm_period180}6m" # Adds "_suffix" to each new column name
      )


tractstorms<-left_join(tracts_stormperiods90, tracts_stormperiods180, by = "BCT2020")

tractstorms <- left_join(tract_overall, tractstorms, by="BCT2020")

#write.csv(tractstorms, "tractstorms.csv")

```

## join to tracts info HU (housing units)
```{r}
#read in tract shapefile with Census pop and housing unit info
tractstorms<-read.csv("tractstorms.csv")
withpopulation <- st_read("nyc_tracts_2020_infoHU_withpop.shp")


tracts<-left_join(withpopulation, tractstorms, by = "BCT2020")

summary(tracts$n_ovrall)

```

## cleaning
```{r}

#calculate average people per unit in tracts across city
# tracts%>%
#   st_drop_geometry()%>%
#    mutate(perunit = totpop/HU_tot)%>%
#    filter(HU_tot>0)%>%
#    summarize(mean = mean(perunit))
# 2.92

#calculate average % rental per tract across city
# tracts%>%
#   st_drop_geometry()%>%
#    mutate(proprenters = HU_rent/HU_tot)%>%
#    filter(HU_tot>0)%>%
#    summarize(mean = mean(proprenters))
# 0.665

tracts<-tracts%>%
  #exclude tracts with 0 housing units
  filter(HU_rent>=100)%>%
  mutate(
    #fill in 0s for tracts with NAs
    n_ovrall = if_else(is.na(n_ovrall), 0, n_ovrall),
    n1000rent = n_ovrall/HU_rent * 1000,
    across(
      .cols = ends_with("3m") | ends_with("6m"),
      .fns  = ~ replace_na(.x, 0)
    ),
         )%>%
  #add rate columns, per 1000 units
  mutate(
    across(
      .cols = ends_with("3m") | ends_with("6m"),
      .fns  = ~ replace_na(.x, 0) / HU_rent *1000,
      .names = "R{col}"
    )
  )

summary(tracts$n_per1000rent)

summary(tracts$RSand6m)

#st_write(tracts, "tracts_withrates2.shp")
```
##GeoDa

NOTE: use neighbors weights file nyc_tracts_2020_infoHU_withpop.gal


##key for LISA results from GeoDa
0 Not significant
1 High-High (red)
2 Low-Low (dark blue)
3 Low-High (light blue)
4 High-Low (pink)

## read in GeoDa LISA results
```{r}
tractsLISA<-st_read("LISAmaps_GeoDa/LISAresults_7-25-26.shp")

LISAresults<- tractsLISA%>%
  st_drop_geometry()

```


## read in shoreline clipped file
```{r}
tractsfull<-st_read("nyctracts2020shorelineclipped/nyct2020.shp")

sf::st_crs(tractsfull)

ggplot(tractsfull)+
  geom_sf()
```

##read borough boundary file
```{r}
boroughboundaries<-st_read("boroughboundaries/boroughboundaries.shp")

boroughboundaries <- st_transform(boroughboundaries, st_crs(tractsfull))

ggplot(boroughboundaries)+
  geom_sf()
```


## join to tract shoreline clipped file
```{r}
tractsfullLISA<-left_join(tractsfull, LISAresults, by="GEOID")
```

## map LISA
```{r}
tractsfullLISA<-tractsfullLISA%>%
    mutate(
    across(
      starts_with("C_"),   # adjust if your columns have different naming
      ~ case_when(
        .x == 0 ~ "Not sig.",
        .x == 1 ~ "High-High",
        .x == 2 ~ "Low-Low",
        .x == 3 ~ "Low-High",
        .x == 4 ~ "High-Low",
        TRUE ~ NA_character_
      ),
      .names = "{.col}"   # keep original & create new labeled vars
    )
  )

# Define colors similar to GeoDa defaults
lisa_colors <- c(
  "Not sig." = "lightgrey",
  "High-High" = "red",
  "Low-Low" = "blue",
  "Low-High" = "lightblue",
  "High-Low" = "pink"
)

# Plot LISA cluster map
LISA_overall<-ggplot(tractsfullLISA) +
  geom_sf(aes(fill = C_overall), color = NA, size = 0.2) +
  geom_sf(data=boroughboundaries, fill=NA,color="black",size=0.5)+
  scale_fill_manual(values = lisa_colors, na.value = "#aaa") +
  theme_void(base_size = 20) +
  labs(fill = "LISA Cluster \nType \nI=0.635")+
  theme(legend.position = c(0.1, 0.95), # Position the legend's reference point
  legend.justification = c("left", "top") # Align the legend's top-left corner to that point
  )
```
## LISA for all storms
```{r}

tracts_LISAstormperiods <- tractsfullLISA %>%
  pivot_longer(names_to = "storm", 
               cols = matches("C_(Ida|Sand|Ophe|Iren)(3m)"), #can add |6m to also do 6m 
               values_to = "cat")%>%
  select(GEOID, storm, cat)%>%
  mutate(storm = case_when(storm=="C_Ida3m" ~"Ida \nI=0.289",
                           storm=="C_Iren3m" ~"Irene \nI=0.181",
                           storm=="C_Ophe3m" ~"Ophelia \nI=0.274",
                           storm=="C_Sand3m" ~"Sandy \nI=0.256",),
         storm = factor(storm, levels=c("Irene \nI=0.181","Sandy \nI=0.256","Ida \nI=0.289","Ophelia \nI=0.274")))

stormLISAmap<-ggplot(tracts_LISAstormperiods) +
  geom_sf(aes(fill = cat), color = NA, size = 0.2) +
  scale_fill_manual(values = lisa_colors, na.value = "#aaa") +
  theme_void(base_size = 18) +
  facet_wrap(~storm)+
  labs(fill = "")+
  theme(strip.text = element_text(size = 18, hjust = 0.1, vjust = -4.5),strip.background = element_blank(), strip.clip = "off")
```



## map rates 
```{r}
tractstorms<- read.csv("tractstorms.csv")

withpopulation <- st_read("nyc_tracts_2020_infoHU_withpop.shp")

withpopulation <- withpopulation%>%
    filter(HU_tot>0)


tracts<-left_join(withpopulation, tractstorms, by = "BCT2020")


tracts_tojoin<-tracts%>%
  st_drop_geometry()

mapping <- left_join(tractsfull, tracts_tojoin, by="GEOID")
```

## map rates overall
```{r}
tracts_overall <- mapping %>%
  pivot_longer(names_to = "storm", 
               cols = n_ovrall, #can add |6m to also do 6m 
               values_to = "rate")%>%
  select(GEOID, storm, rate)

summary(tracts_overall$rate)

breaks<-c(0,50,100,200,1017)

tracts_overall<-tracts_overall%>%
  mutate(rate_breaks=cut(rate, breaks = breaks, include.lowest = TRUE, dig.lab = 10), 
         #dig.lab makes numbers whole numbers not scientific notation
         storm = case_when(storm=="n1000rent" ~"Overall")) 
  
overallratesmap<-ggplot(tracts_overall) +
  geom_sf(aes(fill=rate_breaks), color = NA)+ #"color" sets the border color
  geom_sf(data=boroughboundaries, fill=NA,color="black",size=0.5)+
  geom_sf_label(data=boroughboundaries,
    aes(label = boroname),           # Map the name column to the label aesthetic
    size = 4,                          # Adjust label size
    color = "black",                   # Adjust label color
    fill = "white",                    # Background color for geom_sf_label
    alpha=0.7,
    fun.geometry = sf::st_point_on_surface # Optional: ensures labels are inside irregular polygons
  )+
  theme_void(base_size = 20) + #removes axes
  scale_fill_manual(values =c("#dcf7e4","#7BBD8A","#2E733C","#061C0B"),
    na.value = "grey70")+
  labs(fill = "Complaints \nper 1000 \nrental units")+
  theme(strip.text = element_text(size = 22), strip.clip = "off",
  legend.position = c(0.1, 0.95), # Position the legend's reference point
  legend.justification = c("left", "top") # Align the legend's top-left corner to that point
)

```

##map rates storm3m
```{r}
tracts_stormperiods <- mapping %>%
  pivot_longer(names_to = "storm", 
               cols = matches("(Ida|Sand|Ophe|Iren)(3m)"), #can add |6m to also do 6m 
               values_to = "rate")%>%
  select(GEOID, storm, rate)

summary(tracts_stormperiods$rate)

breaks<-c(0,2,5,10,134)

tracts_stormperiods<-tracts_stormperiods%>%
  mutate(rate_breaks=cut(rate, breaks = breaks, include.lowest = TRUE),
         storm = case_when(storm=="Ida3m" ~"Ida",
                           storm=="Iren3m" ~"Irene",
                           storm=="Ophe3m" ~"Ophelia",
                           storm=="Sand3m" ~"Sandy",),
         storm = factor(storm, levels=c("Irene","Sandy","Ida","Ophelia")))
  
stormratesmap<-ggplot(tracts_stormperiods) +
  geom_sf(aes(fill=rate_breaks), color = NA)+ #"color" sets the border color
  theme_void(base_size = 18) + #removes axes
  scale_fill_manual(values =c("#dcf7e4","#7BBD8A","#2E733C","#061C0B"),
    na.value = "grey70")+
  facet_wrap(~storm)+
  labs(fill = "")+
  theme(strip.text = element_text(size = 18, hjust = 0.1, vjust = -4.5), strip.clip = "off")

stormratesmap
```

# LISA 3m vs 6m sensitivity
```{r}

```


## Export Figure 2
```{r}
library(cowplot)
combined_plot <- plot_grid(
                           #list objects Left -> Right, Top row -> Bottom row
                           overallratesmap, LISA_overall, stormratesmap, stormLISAmap,
                           rel_widths = c(1,1), rel_heights = c(1,1), #specify relative widths and heights if applicable
                           ncol = 2, nrow=2 #specify number of rows/columns if applicable
                           )

# Save the combined plot
save_plot("mold_figure2_grid_7-25-26.png", combined_plot, base_height = 10, base_asp = 1.2, dpi=600)
```



#Figure S1
```{r}
#Read in data
#compute day of week
dailycomparison<-dailycomparison%>%
  mutate(dayofweek = wday(day, label = TRUE), )

#check day of week
table(dailycomparison$dayofweek) #should be equal if full weeks used

#count complaints per day of week
weekdaycount<-dailycomparison%>%
  group_by(dayofweek)%>%
  summarize(Overall = sum(dailyn_overall),
            Mold = sum(dailyn_mold, na.rm=T), #four days had 0 mold complaints (potential system issue 4/12/24-4/14/24?)
            )%>%
  ungroup()

#pivot long for faceted graph
weekdaycount_long <-weekdaycount%>%
  pivot_longer(cols = c(Overall,Mold), names_to = "call_type", values_to = c("Count"))%>%
  mutate(call_type=factor(call_type, levels = c("Overall","Mold")))

weekdaycount_long<-weekdaycount_long%>%
  mutate(calls_per_day= Count/791)

figureS1<-ggplot(weekdaycount_long, aes(x=dayofweek, y=calls_per_day))+
  geom_col()+
  facet_grid(call_type~.,scales = "free_y")+
  scale_y_continuous(
    labels = scales::comma
  )+
  labs(x="Day of week", y="Average calls per day")+
  theme_classic(base_size = 20)


ggsave("mold_figureS1.png", figureS1, width = 10, height = 8, dpi = 300)
```

